---
title: "COVID-19 Vaccination Analysis"
author: "Yash Anand Rajpal, Advait Bhargav, Glit Hanpanitkitkan"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
- \usepackage{graphicx}
- \usepackage{booktabs} 
- \usepackage{caption}
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
library(knitr)
library(tidyverse)
library(LearnBayes)
library(ggplot2)
library(HDInterval)
library(fastR2)
knitr::opts_chunk$set(fig.path = "figures/")  
```

# Abstract
This paper evaluates the efficacy of the Pfizer-BioNTech BNT162b2 COVID-19 vaccine using data from its Phase III clinical trial, in which 8 out of 17,411 vaccinated participants and 162 out of 17,511 placebo participants developed confirmed COVID-19 infections. We adopt a conditional binomial model, modeling the number of vaccine-group cases among all 170 confirmed cases as $T \sim \text{Binom}(170, \pi)$, and reparametrize in terms of vaccine efficacy $\psi = (1 - 2\pi)/(1 - \pi)$. Using likelihood inference, we obtain a maximum likelihood estimate of $\hat{\psi}^{mle} = 95.1\%$ with a 95\% confidence interval of $[91.6\%, 98.6\%]$, and a likelihood ratio test rejects the FDA's minimum efficacy threshold of 30\% ($W = 121.6$, $p \approx 0$). Under a Bayesian framework with a skeptical $\text{Beta}(43.04, 43.04)$ prior centered at $\pi = 0.5$ (corresponding to zero efficacy), we obtain a posterior median efficacy of $75.2\%$ with a 95\% credible interval of $[66.6\%, 82.0\%]$ and a Bayesian p-value of $1.94 \times 10^{-13}$. Both frequentist and Bayesian analyses provide overwhelming evidence that BNT162b2 far exceeds the FDA's 30\% efficacy threshold.

# Keywords
*Vaccine Efficacy*, *Bayesian Inference*, *Likelihood Inference*, *Likelihood Ratio Test*

\newpage

# Introduction / Background
The emergence of COVID-19 in late 2019 and its rapid global spread in 2020 resulted in a crisis affecting healthcare, economics, and daily life worldwide, forcing governments to implement unprecedented public health measures. By the end of 2020, countries such as the United States had seen over a million cases of COVID. This meant the race to find a reliable vaccine was on, and it became a primary goal for pharmaceuticals and governments worldwide in order to control the pandemic. Pfizer and BioNTech developed BNT162b2, an mRNA-based vaccine.

The clinical trial was a placebo-controlled, observer-blinded study that enrolled 43,548 participants aged 16 years or older. The participants were randomly assigned in a 1:1 ratio to receive either two doses of the BNT162b2 vaccine or a placebo, both administered 21 days apart. The primary endpoint was laboratory-confirmed COVID-19 occurring at least 7 days after the second dose (Polack et al., 2020).

The results of the study are summarized^[The total number of people is less than the one in the study because some people left the study or were not reached out to after.] below.

```{r label= "vars", eval=TRUE, echo=FALSE}
vacc_cas <- 8
plac_cas <- 162
n_vacc <- 17411
n_plac <- 17511
tot_cas <- vacc_cas + plac_cas
n_tot <- n_vacc + n_plac
```

\begin{minipage}{0.48\textwidth}
```{r label="table1", eval=TRUE, echo=FALSE}
table_df <- data.frame(
  Group = c("BNT162b2 (Vaccine)", "Placebo", "Total"),
  Cases = c(vacc_cas, plac_cas, tot_cas),
  No_Cases = c(n_vacc - vacc_cas, n_plac - plac_cas, n_tot - tot_cas),
  Total = c(n_vacc, n_plac, n_tot)
)
kable(table_df,
      format = "latex",
      booktabs = TRUE,
      col.names = c("Group", "Infection", "No Infection", "Total"),
      align = "lccc")
```
\captionof{table}{Contingency Table}
\end{minipage}
\hfill
\begin{minipage}{0.48\textwidth}
```{r label="histogram", eval=TRUE, echo=FALSE, fig.width=5, fig.height=4, fig.env='', out.width="\\linewidth"}
bar_data <- data.frame(
  Group = c("BNT162b2", "Placebo"),
  Rate = c(vacc_cas / n_vacc * 100, plac_cas / n_plac * 100)
)
ggplot(bar_data, aes(x = Group, y = Rate, fill = Group)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_fill_manual(values = c("BNT162b2" = "steelblue", "Placebo" = "salmon")) +
  labs(x = "Group", y = "Infection Rate (%)",
       title = "COVID-19 Infection Rates by Treatment Group") +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_text(aes(label = paste0(round(Rate, 3), "%")), vjust = -0.25, size = 4)
```
\end{minipage}

From the table and plot above, we can see that the estimated infection rates are very different. For the vaccinated group, we have: $\hat\pi_v = \frac{8}{17411} = 0.046\%$ and for the placebo group, we have $\hat \pi_p = \frac{162}{17511} = 0.925\%$. 

\newpage
As the probabilities are so small, we can look at the relative risk which can be calculated as:

$$
\widehat{RR} = \frac{\hat{\pi}_v}{\hat{\pi}_p} = \frac{0.046\%}{0.925\%} \approx 0.0497
$$

The risk of COVID-19 in the vaccine group is roughly $5\%$ that of placebo group, so the vaccine decreases the risk of infection by 95\% -- the vaccine "efficacy".


$$\text{vaccine efficacy} = 1 - \widehat{RR} = 1 - \frac{\hat{\pi}_v}{\hat{\pi}_p}$$

The US FDA requires that a new vaccine demonstrates at least $30\%$ efficacy for approval. Polack et al. (2020) reports a vaccine efficacy of $95.0\%$ with a $95\%$ Bayesian interval of $[90.3\%,\; 97.6\%]$ using a beta binomial model. In this paper, we extend their analysis using both frequentist and Bayesian methods to estimate the efficacy $\psi$ and test the hypothesis $H_0 : \psi = 0.30$ vs $H_1 : \psi > 0.30$.

# Statistical Methods

Let $X$ denote the number of COVID-19 cases in the vaccine group among the $n_1 = 17,411$ participants vaccinated with BNT162b2, and let $Y$ denote the number of cases in the placebo group among the $n_2 = 17,511$ placebo participants. We assume $X\sim \text{Binom}(n_1,\pi_v),\; Y\sim \text{Binom}(n_2, \pi_p)$, where $\pi_v$ and $\pi_p$ are the true infection probabilities in the vaccine and placebo groups, and $X, Y$ are independent.

The parameter of interest is the vaccine efficacy:
$$\psi  = 1 -\frac{\pi_v}{\pi_p} = \frac{\pi_p-\pi_v}{\pi_p}$$
Rather than working with two binomial random variables, we adopt a single binomial model. Let $T$ denote the number of vaccine-group cases among the $n=170$ total confirmed COVID-19 cases, conditional on $X+Y=170$. The distribution of $T$ can be shown to be well approximated by a binomial distribution (proof in Appendix).

Then
$$T\sim Binom(n=170, \pi)$$
where $\pi = P(\text{vaccinated} \mid \text{infected})$ and can be shown to be equal to
$$\pi = \frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p} \approx \frac{\pi_v}{\pi_v + \pi_p}$$

The vaccine efficacy $\psi$ can be written in terms of $\pi$ as:
$$\psi = \frac{1-2\pi}{1-\pi}, \quad -\infty < \psi < 1$$

We are interested in an interval estimate for this parameter ($\psi$).

## Likelihood Inference

Given the observed $t = 8$ vaccine-group cases, the likelihood function under $T \sim \text{Binom}(170, \pi)$ is:

$$
L(\pi) = \binom{170}{8}\pi^8 (1 - \pi)^{162}
$$

Since $\psi = \cfrac{1-2\pi}{1-\pi}$, solving for $\pi$ and substituting into the likelihood function yields:
\begin{align*}
  \psi &= \frac{1-2\pi}{1-\pi}, \quad -\infty < \psi < 1 \\
  \psi (1 - \pi) &= 1 - 2 \pi \\
  \psi -\psi\pi  &= 1 - 2 \pi \\
  2\pi - \psi\pi &= 1 - \psi \\
  \pi (2 - \psi) &= 1 - \psi \\
  \pi &= \frac{1 - \psi}{2 - \psi}\\
\end{align*}
\begin{align*}
L^{\ast} (\psi) &= \binom{170}{8} \left(\frac{1 - \psi}{2 - \psi}\right)^8 \left(1 - (\frac{1 - \psi}{2 - \psi})\right)^{162} \quad  \text{after substituting for }  \pi \\
                &= \binom{170}{8} \left(\frac{1 - \psi}{2 - \psi}\right)^8 \left(\frac{1}{2 - \psi}\right)^{162}
\end{align*}
Taking the log of the likelihood function:
\begin{align*}
 \ell = \ln [L^\ast (\psi)] &= \ln\left(\binom{170}{8}\right) + 8 \ln\left(\frac{1 - \psi}{2 - \psi}\right) + 162 \ln (\frac{1}{2 - \psi}) \\
                            &= \ln\left(\binom{170}{8}\right) + 8 \ln(1 - \psi) - 8\ln(2 - \psi) - 162 \ln (2 - \psi) \\
                            &= \ln\left(\binom{170}{8}\right) + 8 \ln(1 - \psi) -170 \ln (2 - \psi)
\end{align*}
Now we take the derivative of the log likelihood function $\ell$ with respect to $\psi$ and set it equal to 0: 
\begin{align*}
\ell' = \frac{-8}{1 - \psi} + \frac{170}{2 - \psi}&\\
\frac{-8}{1 - \psi} + \frac{170}{2 - \psi} &= 0 \\
-8(2 - \psi) + 170(1 - \psi) &= 0 \\
-16 + 8 \psi + 170 -170\psi &= 0 \\
\hat \psi_0 ^{\text{mle}} = \frac{154}{162}
\end{align*}

We verify this is a global maximum via the second derivative test: $$\ell'' = \frac{-8}{(1 - \psi)^2} + \frac{170}{(2 - \psi)^2}$$
Evaluating at the MLE, we have $\ell''(\hat{\psi}^{mle}) = -3126.12 < 0$, 
which confirms $\ell''(\hat{\psi}^{mle}) < 0$ and hence 
$\hat{\psi}^{mle} = \frac{154}{162}$ is a global maximum.

Assuming the regularity conditions are satisfied, 
$$
\hat \psi^{mle}_0 \approx Norm \left(\psi_0, \frac{1}{\sqrt{I(\psi_0)}}\right)
$$
where $I(\psi_0) = n E[- \ell''(\psi)]$. We observe $t=8$, so
\begin{align*}
I({\psi}_0) &= -\ell''(\hat{\psi}_0^{mle}) \\
                &= -\left( \frac{-8}{(1-\hat{\psi}_0^{mle})^2} + \frac{170}{(2-\hat{\psi}_0^{mle})^2} \right) \\
                &= \frac{8}{(1-\hat{\psi}_0^{mle})^2} - \frac{170}{(2-\hat{\psi}_0^{mle})^2} \\
                &= \frac{8}{\left(1-\frac{154}{162}\right)^2} - \frac{170}{\left(2-\frac{154}{162}\right)^2} \\
                &= 3126.1235
\end{align*}
$$
\hat \psi^{mle}_0 \approx Norm \left(\psi_0, \sqrt{\frac{1}{3126.1235}}\right)
$$
We now construct a $95\%$ large-sample confidence interval for $\psi$ using the formula:

$$
\hat \psi_0^\text{mle} \pm z_\frac{\alpha}{2} \widehat{SE} = \hat \psi_0^\text{mle} \pm z_\frac{\alpha}{2} \frac{1}{\sqrt{I(\hat \psi_0^\text{mle})}}
$$

```{r label = "confidence interval", eval = T, echo = FALSE}
mle = 154/162
z = 1.96
se = 1/ sqrt(3126.1235)

lower = mle - z*se
upper = mle + z*se
```


```{r label = "bootstrap", eval = TRUE, echo = FALSE}
set.seed(394)

B <- 1000   
n <- 170
Tobs <- 8

# MLE
psi_mle <- (n - 2*Tobs)/(n - Tobs)

# Converting it to pi
pi_hat <- (1 - psi_mle)/(2 - psi_mle)

# Bootstrap
psi_boot <- replicate(n = B, expr = {
      Tb <- rbinom(1, n, pi_hat)
      (n - 2*Tb)/(n - Tb)}
    )
```

We also construct a parametric bootstrap percentile interval by resampling $T^* \sim \text{Binom}(170, \hat{\pi})$ for $B = 1000$ replicates.

```{r label = "Second order approximation plot", eval=T, echo=F, warning=F, fig.height=3}
loglik.psi <- function(psi){
  ifelse((psi >= 1), NA,
         log(choose(170,8)) + 8*log(1-psi) - 170*log(2-psi))
}

ml.psi <- maxLik2(loglik = loglik.psi, start = 0.9)

plot(ml.psi) +
  labs(title = "Second order approximation",
       subtitle = "n = 170, t = 8",
       x = expression(psi))
```

The plot shows that the log-likelihood is extremely well approximated by the second-order Taylor expansion about the MLE, confirming that the normal approximation to the sampling distribution of $\hat\psi^{mle}$ is appropriate.

Now, we test
$$
H_0 : \psi = 0.30\qquad \text{vs} \qquad H_1 : \psi > 0.30$$

The likelihood ratio test statistic is based on
$$
W = 2\ln \Lambda  \quad \text{where} \quad \Lambda = \frac{L(\hat \psi_0^\text{mle})}{L(\psi_0^\text{null})}
$$
$$
W = 2\left[\ell(\hat\psi_0^\text{mle}) - \ell(\psi_0^\text{null})\right].
$$
Under $H_0$, the test statistic satisfies $W \sim \chi^2_1$ as we are only testing one parameter, and the p-value is $P(\chi^2_1 \ge W)$.
```{r label = "pval", eval = TRUE, echo = FALSE}
W <- 2*((8*log(8 / (162 * 0.7)) - 170*log(170/ (162* 1.7)))) 
p_val <- pchisq(W, df = 1, lower.tail = FALSE)
```

We also estimate the p-value empirically by simulating $B = 1000$ samples under $H_0$ and computing the proportion of simulated test statistics that exceed the observed $W$.

```{r label = "empirical pval", eval = TRUE, echo = FALSE}
psi0 = 0.3
pi0 = (1 - psi0) / (2 - psi0)

# observed LRT statistic
W = 2*(loglik.psi(psi_mle) - loglik.psi(psi0))

w = numeric(B)

# Bootstrap
for(b in 1:B){

  Tb = rbinom(1,n,pi0)

  psi_mle = (n - 2*Tb)/(n - Tb)

  log_hat = loglik.psi(psi_mle)
  log_0   = loglik.psi(psi0)

  w[b] = 2*(log_hat - log_0)
}

# empirical p-value
p_val_b = mean(w >= W)
```

```{r label = "simulated Ws under null histogram", include = FALSE}
hist(w, breaks = 20, 
     xlab = "W", 
     ylab = "Frequency", 
     main = "Histogram of simulated W* values under null", 
     col = "lightblue",
     xlim = c(-60, 125))
```

## Bayesian Inference

In a Bayesian framework, we specify a prior distribution for $\pi$. A common choice for the binomial model is the Beta prior: $$g(\pi) = Beta(a, b)$$
We chose a skeptical prior that reflects uncertainty about whether the vaccine is effective. This allows the data to play a larger role in determining the posterior distribution and helps prevent incorrectly concluding that the vaccine is effective when it may not be.

Coming up with two prior beliefs about $\psi$:

\begin{enumerate}
  \item For $\psi \ge 0.30$, we have
  $$
  \frac{1 - 2\pi}{1 - \pi} \ge 0.30.
  $$
  so
  $$
  \pi \le \frac{0.70}{1.70} = \frac{7}{17}.
  $$
  Therefore,
  $$
  P\left(\pi \le \frac{7}{17}\right) = 0.05.
  $$

  \item For $\psi \ge 0$, we have
  $$
  \frac{1 - 2\pi}{1 - \pi} \ge 0.
  $$
  Since $1-\pi > 0$ for $0 < \pi < 1$, this is equivalent to
  $$
  1 - 2\pi \ge 0.
  $$
  Hence,
  $$
  \pi \le \frac{1}{2}.
  $$
  Therefore,
  $$
  P\left(\pi \le \frac{1}{2}\right) = 0.50.
  $$
\end{enumerate}

So we want a $\text{Beta}(a, b)$ prior for $\pi$ where the median is $\frac{1}{2}$ and the 5th percentile is $\frac{7}{17}$, or $P(\pi \le \frac{7}{17}) = 0.05$. Using the `beta.select` function from the `LearnBayes` library to find the beta prior, we obtain $a = 43.04$ and $b= 43.04$.

```{r label = "beta coeffs", eval=TRUE, echo=FALSE}
b_coeff = beta.select(quantile1 = list(p = 0.05, x = 7/17),
                      quantile2 = list(p=0.5, x = 0.5))
```

With the skeptical $Beta(43.04, 43.04)$, we have the beta prior for $\pi$:

```{r label = "beta prior plot", eval = TRUE, echo=FALSE, fig.height=3.25, fig.align='center'}
x_values = seq(0, 1, length.out = 1000)
alpha = b_coeff[1]
beta = b_coeff[2]
density_values = dbeta(x_values, shape1 = alpha, shape2 = beta)

plot(x_values, density_values, type = 'l',
     xlab = expression(pi), ylab = "PDF",
     main = paste("Beta(", alpha, ",", beta, ") prior for vaccine data", sep = ""))
```

The Beta distribution is conjugate to the binomial and so the posterior must also be a Beta distribution.
\begin{align*}
f(T\mid\pi) &\sim Binom(170, \pi) \text{ as the data model,} \\
g(\pi) &= Beta(43.04, 43.04) \text{ as the prior} \\
h(\pi\mid t) &= Beta(43.04 + t, 170 + 43.04 - t) \text{ as the posterior} \\
\text{where t = 8, } h(\pi\mid 8) &= Beta(43.04 + 8, 170 + 43.04 - 8) \\
\text{thus } h(\pi \mid 8) &= Beta(51.04, 205.04)
\end{align*}

```{r label = "prior vs post plot", eval = TRUE, echo = FALSE, fig.height=4, fig.align='center'}
x_values <- seq(0, 1, length.out = 1000)
n = 170
t = 8

# posterior parameters
alpha_post <- alpha + t
beta_post  <- beta + (n - t)

# densities
prior_density <- dbeta(x_values, shape1 = alpha, shape2 = beta)
post_density  <- dbeta(x_values, shape1 = alpha_post, shape2 = beta_post)

plot_data <- data.frame(
  pi = rep(x_values, 2),
  density = c(prior_density, post_density),
  distribution = rep(c("prior", "posterior"), each = length(x_values))
)

# plot
ggplot(plot_data, aes(x = pi, y = density, color = distribution)) +
  geom_line(linewidth = 0.5) +
  labs(
    title = "Beta prior and posterior for vaccine data",
    x = expression(pi),
    y = "PDF",
    color = "distribution"
  ) +
  theme_minimal()
```

```{r label = "posterior median", eval = TRUE, echo = FALSE, results='hide'}
posterior_median_pi = qbeta(0.5, shape1 = 51.04, shape2 = 205.04)
cat("Posterior Median: ", round(posterior_median_pi, 4))
cat("95% interval for pi (equal tails): (15.28, 25.03)")
```

```{r label = "posterior distribution", eval = TRUE, echo = FALSE, results='hide'}
qbeta(p = c(0.025, 0.975), shape1 = 51.04, shape2 = 205.04)
```

We also compute the $95\%$ equal-tail credible interval and highest posterior density interval (HPDI) for $\pi$.

```{r label = "hdi interval", eval = TRUE, echo = FALSE, results='hide'}
posterior_interval = hdi(qbeta, credMass=0.95, shape1 = 51.04,
shape2 = 205.04)

cat("95% HPDI: ", round(posterior_interval, 4))
```

To convert posterior summaries from $\pi$ to the vaccine efficacy scale $\psi$, we use the transformation $\psi = (1-2\pi)/(1-\pi)$. Since $\psi$ is a decreasing function of $\pi$, the lower bound of an interval for $\pi$ maps to the upper bound for $\psi$, and vice versa.

### Bayesian P-value

The Bayesian p-value is defined as the posterior probability that the null hypothesis is true:
\begin{align*}
\text{Bayes P-value} &: P(H_0 \text{ is TRUE }\mid X) \\
&= P(\psi \le \psi_0^{(null)}\mid X=x)
\end{align*}
In our setting, this becomes $P(\psi \le 0.30 \mid t=8) = P\left(\pi \ge \frac{0.7}{1.7}\mid t=8\right)$, which we compute from the posterior Beta distribution.

```{r label = "bayesian pval", eval = TRUE, echo = FALSE, results='hide'}
pbeta(0.7/1.7, shape1 = 51.04, shape2 = 205.04, lower.tail = FALSE)
```

# Results

## Likelihood Results

The maximum likelihood estimate of vaccine efficacy is $\hat{\psi}^{mle} = \frac{154}{162} \approx 95.1\%$. Table 2 summarizes the frequentist interval estimates for $\psi$.

\begin{table}[htbp]
\centering
\caption{Frequentist Interval Estimates for Vaccine Efficacy ($\psi$)}
\begin{tabular}{lc}
\toprule
Method & 95\% Interval \\
\midrule
Large-Sample CI & [91.6\%, 98.6\%] \\
Bootstrap Percentile CI & [91.0\%, 98.2\%] \\
\bottomrule
\end{tabular}
\end{table}

Both intervals place the vaccine efficacy well above $90\%$.

The likelihood ratio test of $H_0: \psi = 0.30$ versus $H_1: \psi > 0.30$ yields a test statistic of:

$$W = 2\left[\ell\!\left(\tfrac{154}{162}\right) - \ell(0.3)\right] = 121.6$$

with a p-value of $P(\chi^2_1 \ge 121.6) = 2 \times 10^{-28} \approx 0$, providing significant evidence to reject $H_0$. The empirical simulation confirms this as none of the $B=1000$ simulated test statistics under $H_0$ exceeded the observed $W$, yielding an estimated p-value of $< 0.001$.

```{r label = "results histogram", eval = TRUE, echo = FALSE, fig.height=4, fig.align='center'}
hist(w, breaks = 20,
     xlab = "W",
     ylab = "Frequency",
     main = "Histogram of simulated W* values under null",
     col = "lightblue",
     xlim = c(-60, 125))
```

## Bayesian Results

\begin{table}[htbp]
\centering
\caption{Bayesian Posterior Summaries}
\begin{tabular}{lcc}
\toprule
Quantity & $\pi$ & Vaccine Efficacy ($\psi$) \\
\midrule
Posterior Median & 0.1985 & 75.2\% \\
95\% Equal-Tail CI & [0.1528, 0.2503] & [66.6\%, 82\%] \\
95\% HPDI & [0.1513, 0.2486] & [66.9\%, 82.2\%] \\
\bottomrule
\end{tabular}
\end{table}

The posterior probability that the vaccine does not meet the FDA's $30\%$ efficacy requirement is:

$$P(\psi \le 0.30\mid t=8) = P\biggr(\pi \ge \frac{0.7}{1.7} \mid t=8\biggr) = 1.939\times10^{-13}$$

Thus, there is significant evidence that the vaccine exceeds the $30\%$ efficacy requirement set by the FDA.

# Discussion / Conclusion

Both the frequentist and Bayesian analyses provide strong evidence that the Pfizer-BioNTech BNT162b2 vaccine is highly effective and far exceeds the FDA's minimum $30\%$ efficacy threshold. The frequentist MLE of $\hat{\psi} = 95.1\%$ with a $95\%$ confidence interval of $[91.6\%, 98.6\%]$ is consistent with the likelihood ratio test ($W = 121.6$, $p \approx 0$), which rejects $H_0: \psi = 0.30$. The Bayesian analysis, using a deliberately skeptical $\text{Beta}(43.04, 43.04)$ prior (shown in the plot with the prior and posterior distributions), yields a lower posterior median efficacy of $75.2\%$ with a $95\%$ equal-tail interval of $[66.6\%, 82.0\%]$. Despite the conservative prior, the Bayesian p-value of $1.94 \times 10^{-13}$ still overwhelmingly rejects the $30\%$ threshold.

Pfizer reported a vaccine efficacy of $95.0\%$ with a $95\%$ Bayesian credible interval of $[90.3\%, 97.6\%]$ and a posterior probability exceeding $0.9999$ that efficacy surpassed $30\%$ using a Beta$(0.700102, 1)$ prior. Our frequentist MLE of $95.1\%$ is nearly identical to their point estimate, however, our Bayesian results differ substantially since our prior is far more conservative than Pfizer's (which is nearly flat and allows the data to dominate the posterior).

```{r label = "prior comparison", eval = TRUE, echo = FALSE, fig.height=4, fig.align='center'}
x_values <- seq(0, 1, length.out = 1000)

alpha_pfizer <- 0.700102
beta_pfizer <- 1
alpha_pfizer_post <- alpha_pfizer + 8
beta_pfizer_post <- beta_pfizer + 162

alpha_ours <- 43.04
beta_ours <- 43.04
alpha_ours_post <- alpha_ours + 8
beta_ours_post <- beta_ours + 162

plot_data <- data.frame(
  pi = rep(x_values, 4),
  density = c(
    dbeta(x_values, alpha_pfizer, beta_pfizer),
    dbeta(x_values, alpha_pfizer_post, beta_pfizer_post),
    dbeta(x_values, alpha_ours, beta_ours),
    dbeta(x_values, alpha_ours_post, beta_ours_post)
  ),
  curve = rep(c("Pfizer prior: (0.70, 1)",
                "Pfizer posterior: (8.70, 163)",
                "Our prior: (43.04, 43.04)",
                "Our posterior: (51.04, 205.04)"),
              each = length(x_values))
)

ggplot(plot_data, aes(x = pi, y = density, color = curve, linetype = curve)) +
  geom_line(linewidth = 0.6) +
  scale_linetype_manual(values = c("solid", "solid", "dashed", "dashed")) +
  labs(
    title = "Comparison of Pfizer and Skeptical Priors/Posteriors",
    x = expression(pi),
    y = "PDF",
    color = "", linetype = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.text = element_text(size = 7))

```

This explains why Pfizer's Bayesian interval closely matches the frequentist results, while ours is pulled toward lower efficacy.

Both methods we employed have strengths and limitations. The frequentist likelihood approach provides inference and intervals without needing a prior, but relies on large-sample normal approximations that may be less accurate for small counts such as the $8$ vaccine cases observed. The Bayesian approach allows incorporation of prior beliefs and also gives us direct probability statements about the parameter, but results are sensitive to the choice of prior (as shown by the gap between our posterior median and Pfizer's). Despite these trade-offs, both approaches agree that BNT162b2 far exceeds the regulatory minimum for vaccine efficacy.

\newpage

# Bibliography

Polack, F. P., Thomas, S. J., Kitchin, N., Absalon, J., Gurtman, A., Lockhart, S., Perez, J. L., P\'{e}rez Marc, G., Moreira, E. D., Zerbini, C., Bailey, R., Swanson, K. A., Roychoudhury, S., Koury, K., Li, P., Kalina, W. V., Cooper, D., Frenck, R. W., Hammitt, L. L., ... Gruber, W. C. (2020). Safety and Efficacy of the BNT162b2 mRNA Covid-19 Vaccine. *New England Journal of Medicine*, 383(27), 2603--2615.


# Appendix
## Code

Initializing variables
```{r ref.label = "vars", eval=FALSE}

```

Contingency table
```{r ref.label = "table1", eval=FALSE}

```

Proportions histogram
```{r ref.label = "histogram", eval=FALSE}

```

Large sample confidence interval    
```{r ref.label = "confidence interval", eval=FALSE}

```

Second order Taylor approximation plot
```{r ref.label = "Second order approximation plot", eval=FALSE}

```

Bootstrap percentile interval
```{r ref.label = "bootstrap", eval=FALSE}

```

P value
```{r ref.label = "pval", eval=FALSE}

```

Empirical P value
```{r ref.label = "empirical pval", eval=FALSE}

```

Simulated Ws under null histogram
```{r ref.label = "simulated Ws under null histogram", eval=FALSE}

```

Producing alpha and beta for beta distribution.
```{r ref.label = "beta coeffs", eval = FALSE}

```

Plotting Beta$(\alpha, \beta)$ prior distribution
```{r ref.label = "beta prior plot", eval = FALSE}

```

Plotting beta prior and posterior distributions
```{r ref.label="prior vs post plot", eval=FALSE}

```

Calculating posterior median
```{r ref.label="posterior median", eval = FALSE}

```

Getting posterior distribution
```{r ref.label = "posterior distribution", eval = FALSE}

```

Highest density interval
```{r ref.label = "hdi interval", eval = FALSE}

```

Bayesian p-value
```{r ref.label="bayesian pval", eval = FALSE}

```

Pfizer prior & posterior plot
```{r ref.label = "pfizer prior posterior", eval = F}

```
## Proofs

Let $X \sim \text{Binom}(n_1, \pi_v)$ be the number of cases in the vaccine group and $Y \sim \text{Binom}(n_2, \pi_p)$ be the number of cases in the placebo group, with $X$ and $Y$ independent. Define $T = X$ and $n = X + Y$. We wish to show that $T \mid X + Y = n$ is approximately $\text{Binom}(n, \pi)$ where $\pi = \frac{n_1 \pi_v}{n_1 \pi_v + n_2 \pi_p}$.

Since $\pi_v$ and $\pi_p$ are both very small, the binomial distributions for $X$ and $Y$ are well approximated by Poisson distributions:
$$X \approx \text{Poisson}(\lambda_1), \quad Y \approx \text{Poisson}(\lambda_2)$$
where $\lambda_1 = n_1 \pi_v$ and $\lambda_2 = n_2 \pi_p$.

By the additive property of independent Poisson random variables:
$$X + Y \approx \text{Poisson}(\lambda_1 + \lambda_2)$$

For $t = 0, 1, \ldots, n$, using the definition of conditional probability:
\begin{align*}
P(X = t \mid X + Y = n) &= \frac{P(X = t,\; Y = n - t)}{P(X + Y = n)} \\
&= \frac{P(X = t) \cdot P(Y = n - t)}{P(X + Y = n)} \\
&= \frac{\dfrac{e^{-\lambda_1} \lambda_1^t}{t!} \cdot \dfrac{e^{-\lambda_2} \lambda_2^{n-t}}{(n-t)!}}{\dfrac{e^{-(\lambda_1+\lambda_2)} (\lambda_1+\lambda_2)^n}{n!}} \\
&= \frac{n!}{t!(n-t)!} \cdot \frac{\lambda_1^t \cdot \lambda_2^{n-t}}{(\lambda_1 + \lambda_2)^n} \\
&= \binom{n}{t} \left(\frac{\lambda_1}{\lambda_1+\lambda_2}\right)^t \left(\frac{\lambda_2}{\lambda_1+\lambda_2}\right)^{n-t}
\end{align*}

This is the probability mass function of a $\text{Binom}\left(n,\; \frac{\lambda_1}{\lambda_1+\lambda_2}\right)$ distribution. Therefore:
$$T \mid X + Y = n \;\sim\; \text{Binom}\left(n,\; \pi\right) \quad \text{where } \pi = \frac{\lambda_1}{\lambda_1 + \lambda_2} = \frac{n_1\pi_v}{n_1\pi_v + n_2\pi_p}$$

In our setting, $n_1 = 17{,}411 \approx n_2 = 17{,}511$, so $\frac{n_1}{n_2} \approx 1$ and the expression simplifies to $\pi \approx \frac{\pi_v}{\pi_v + \pi_p}$.