---
title: "COVID-19 Vaccination Analysis"
author: "Yash Anand Rajpal, Advait Bhargav, Glit Hanpanitkitkan"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
- \usepackage{graphicx}
- \usepackage{booktabs} 
- \usepackage{caption}
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
library(knitr)
library(tidyverse)
library(LearnBayes)
library(ggplot2)
library(HDInterval)
library(fastR2)
knitr::opts_chunk$set(fig.path = "figures/")  
```

# Abstract
Write your abstract here.

# Keywords
*Vaccine Efficacy*, *Bayesian Inference*, *Likelihood Inference*, *Likelihood Ratio Test*

\newpage

# Introduction / Background
The rise of COVID-19 in 2019 resulted in a global crisis affecting healthcare, economics, and daily life across societies worldwide, forcing governments to implement unprecedented public health measures. By the end of 2020, countries such as the United States had seen over a million cases of COVID. This meant the race to find a reliable vaccine was on, and it became a primary goal for pharmaceuticals and governments worldwide in order to control the pandemic. Pfizer and BioNTech developed BNT162b2, an mRNA based vaccine.

The clinical trial was a placebo-controlled, observer-blinded study that enrolled 43,548 participants aged 16 year or older. The participants were randomly assigned in a 1:1 ratio to receive either two doses of the BNT162b2 vaccine or a placebo, both administered 21 days apart. The primary endpoint was laboratory-confirmed COVID-19 occurring at least 7 days after the second dose (Polack et al., 2020).

The results of the study are summarized^[The total number of people is less than the one in the study because some people left the study or were not reached out to after.] below.

```{r label= "vars", eval=TRUE, echo=FALSE}
vacc_cas <- 8
plac_cas <- 162
n_vacc <- 17411
n_plac <- 17511
tot_cas <- vacc_cas + plac_cas
n_tot <- n_vacc + n_plac
```

\begin{minipage}{0.48\textwidth}
```{r label="table1", eval=TRUE, echo=FALSE}
table_df <- data.frame(
  Group = c("BNT162b2 (Vaccine)", "Placebo", "Total"),
  Cases = c(vacc_cas, plac_cas, tot_cas),
  No_Cases = c(n_vacc - vacc_cas, n_plac - plac_cas, n_tot - tot_cas),
  Total = c(n_vacc, n_plac, n_tot)
)
kable(table_df,
      format = "latex",
      booktabs = TRUE,
      col.names = c("Group", "Infection", "No Infection", "Total"),
      align = "lccc")
```
\captionof{table}{Contingency Table}
\end{minipage}
\hfill
\begin{minipage}{0.48\textwidth}
```{r label="histogram", eval=TRUE, echo=FALSE, fig.width=5, fig.height=4, fig.env='', out.width="\\linewidth"}
bar_data <- data.frame(
  Group = c("BNT162b2", "Placebo"),
  Rate = c(vacc_cas / n_vacc * 100, plac_cas / n_plac * 100)
)
ggplot(bar_data, aes(x = Group, y = Rate, fill = Group)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_fill_manual(values = c("BNT162b2" = "steelblue", "Placebo" = "salmon")) +
  labs(x = "Group", y = "Infection Rate (%)",
       title = "COVID-19 Infection Rates by Treatment Group") +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_text(aes(label = paste0(round(Rate, 3), "%")), vjust = -0.25, size = 4)
```
\end{minipage}

From the table and plot above, we can see that the estimated infection rates are very different. For the vaccinated group, we have: $\hat\pi_v = \frac{8}{17411} = 0.046\%$ and for the placebo group, we have $\hat \pi_p = \frac{162}{17511} = 0.925\%$. 

\newpage
As the probabilities are so small, we can look at the relative risk which can be calculated as:

$$
\widehat{RR} = \frac{\hat{\pi}_v}{\hat{\pi}_p} = \frac{0.046\%}{0.925\%} \approx 0.0497
$$

The risk of COVID-19 in the vaccine group is roughly $5\%$ that of placebo group, so the vaccine decreases the risk of infection by 95\% -- the vaccine "efficacy".


$$\text{vaccine efficacy} = 1 - \widehat{RR} = 1 - \frac{\hat{\pi}_v}{\hat{\pi}_p}$$

The US FDA requires that a new vaccine demonstrates at least $30\%$ efficacy for approval. Polack et al. (2020) reports a vaccine efficacy of $95.0\%$ with a $95\%$ Bayesian credible interval of $[90.3\%,\; 97.6\%]$ using a beta binomial model. In this paper, we intend to extend their analysis by using both Frequentist and Bayesian methods to estimate the efficacy $\psi$ and test the hypothesis $H_0 : \psi = 0.30$ vs $H_1 : \psi > 0.30$.

# Statistical Methods

## Model

Let $X$ denote the number of COVID-19 cases in the vaccine group among the $n_1 = 17,411$ vaccinated with BNT162b2 participants, and let Y denote the number of cases in the placebo group among the $n_2 = 17,511$ placebo participants. We assume $X\sim \text{Binom}(n_1,\pi_v),\; Y\sim \text{Binom}(n_2, \pi_p)$, where $\pi_v$ and $\pi_p$ are the true infection probabilities in the vaccine and placebo groups, and $X, Y$ are independent.

The parameter of interest is the vaccine efficacy:
$$\psi  = 1 -\frac{\pi_v}{\pi_p} = \frac{\pi_p-\pi_v}{\pi_p}$$
Rather than working with two binomial random variables, we adopt a single binomial model. Let $T$ denote the number of vaccine-group cases among the $n=170$ total confirmed COVID-19 cases, conditional on $X+Y=170$. The distribution of $T$ can be shown to be well approximated by a binomial distribution.

Then
$$T\sim Binom(n=170, \pi)$$
where $\pi = P(\text{vaccine} \mid \text{case})$ and can be shown to be equal to
$$\pi = \frac{n_1\pi_v}{n_1\pi_v+n_2\pi_p} = \frac{\pi_v}{\pi_v + \pi_p}$$

The vaccine efficacy $\psi$ can be written in terms of $\pi$ as:
$$\psi = \frac{1-2\pi}{1-\pi}, \quad -\infty < \psi < 1$$
\begin{enumerate}
    \item When $\pi = 0$ (no cases in the vaccinated group), the efficacy is perfect ($\psi = 1$).
    
    \item When $\pi = 0.5$ (proportion of cases is equal in both vaccinated and control groups), there is no efficacy: $\psi = 0$.
    
    \item If $\psi < 0$, it indicates there are more infections in the group with the vaccine, so the vaccine is harmful rather than helpful.
\end{enumerate}

From the table, we observe $t=8$ vaccine cases among the $n=170$ total cases. This seems very small, but we should still conduct tests to verify.

We are interested in an interval estimate for this parameter ($\psi$).

## Likelihood Inference

For $t = 8$ vaccine cases, The likelihood function for $\pi$ for $T \sim Binom(n = 170, \pi)$ can be written as follows:

$$
L(\pi) = \binom{170}{8}\pi^8 (1 - \pi)^{162}
$$
Since, $\psi = \cfrac{1-2\pi}{1-\pi},$, by solving for $\pi$ and then plugging it into the likelihood function we get:
\begin{align*}
  \psi &= \frac{1-2\pi}{1-\pi}, \quad -\infty < \psi < 1 \\
  \psi (1 - \pi) &= 1 - 2 \pi \\
  \psi -\psi\pi  &= 1 - 2 \pi \\
  2\pi - \psi\pi &= 1 - \psi \\
  \pi (2 - \psi) &= 1 - \psi \\
  \pi &= \frac{1 - \psi}{2 - \psi}\\
L^{\ast} (\psi) &= \binom{170}{8} \left(\frac{1 - \psi}{2 - \psi}\right)^8 \left(1 - (\frac{1 - \psi}{2 - \psi})\right)^{162} \quad  \text{after substituting for }  \pi \\
                &= \binom{170}{8} \left(\frac{1 - \psi}{2 - \psi}\right)^8 \left(\frac{1}{2 - \psi}\right)^{162}
\end{align*}
Taking the log of the likelihood function:
\begin{align*}
 \ell = \ln [L^\ast (\psi)] &= \ln\left(\binom{170}{8}\right) + 8 \ln\left(\frac{1 - \psi}{2 - \psi}\right) + 162 \ln (\frac{1}{2 - \psi})^{162} \\
                            &= \ln\left(\binom{170}{8}\right) + 8 \ln(1 - \psi) - 8\ln(2 - \psi) - 162 \ln (2 - \psi) \\
                            &= \ln\left(\binom{170}{8}\right) + 8 \ln(1 - \psi) -170 \ln (2 - \psi)
\end{align*}
Now we take the derivative of the log likelihood function $\ell$ with respect to $\psi$ and set it equal to 0: \newpage

$$\ell' = \frac{-8}{1 - \psi} + \frac{170}{2 - \psi}$$
\begin{align*}
\frac{-8}{1 - \psi} + \frac{170}{2 - \psi} &= 0 \\
-8(2 - \psi) + 170(1 - \psi) &= 0 \\
-16 + 8 \psi + 170 -170\psi &= 0 \\
\hat \psi_0 ^{\text{mle}} = \frac{154}{162}
\end{align*}

We can take the 2nd derivative test to check if it is global maximum:

$$\ell'' = \frac{-8}{(1 - \psi)^2} + \frac{170}{(2 - \psi)^2}$$
Evaluating at the MLE, we have $-\ell''(\hat{\psi}^{mle}) = 3126.12 > 0$, 
which confirms $\ell''(\hat{\psi}^{mle}) < 0$ and hence 
$\hat{\psi}^{mle} = \frac{154}{162}$ is a global maximum.

Assuming the regularity conditions are satisfied, 
$$
\hat \psi^{mle}_0 \approx Norm \left(\psi_0, \frac{1}{\sqrt{I(\psi_0)}}\right)
$$
where $I(\psi_0) = n E[- \ell''(\psi)]$. We observe $t=8$, so
\begin{align*}
I({\psi}_0) &= -\ell''(\hat{\psi}_0^{mle}) \\
                &= -\left( \frac{-8}{(1-\hat{\psi}_0^{mle})^2} + \frac{170}{(2-\hat{\psi}_0^{mle})^2} \right) \\
                &= \frac{8}{(1-\hat{\psi}_0^{mle})^2} - \frac{170}{(2-\hat{\psi}_0^{mle})^2} \\
                &= \frac{8}{\left(1-\frac{154}{162}\right)^2} - \frac{170}{\left(2-\frac{154}{162}\right)^2} \\
                &= 3126.1235
\end{align*}
$$
\hat \psi^{mle}_0 \approx Norm \left(\psi_0, \sqrt{\frac{1}{3126.1235}}\right)
$$
And, now we can calculate a $95\%$ large sample confidence interval for $\psi_0$ by using the formula:

$$
\hat \psi_0^\text{mle} \pm z_\frac{\alpha}{2} \widehat{SE} = \hat \psi_0^\text{mle} \pm z_\frac{\alpha}{2} \frac{1}{\sqrt{I(\hat \psi_0^\text{mle})}}
$$

```{r label = "confidence interval", eval = T, echo = FALSE}
mle = 154/162
z = 1.96
se = 1/ sqrt(3126.1235)

lower = mle - z*se
upper = mle + z*se
```

By plugging in the values, we can say that we are $95\%$ confident that the true value for $\psi_0$ lies in the interval $[0.9156, 0.9857]$ 

```{r label = "bootstrap", eval = TRUE, echo = FALSE}
set.seed(394)

B <- 1000   
n <- 170
Tobs <- 8

# MLE
psi_mle <- (n - 2*Tobs)/(n - Tobs)

# Converting it to pi
pi_hat <- (1 - psi_mle)/(2 - psi_mle)

# Bootstrap
psi_boot <- replicate(n = B, expr = {
      Tb <- rbinom(1, n, pi_hat)
      (n - 2*Tb)/(n - Tb)}
    )
```
The parametric bootstrap percentile interval for $\psi_0$ is [$91.03,\;98.20$].

```{r label = "Second order approximation plot", eval=T, echo=F, warning=F, fig.height=4}
loglik.psi <- function(psi){
  ifelse((psi >= 1), NA,
         log(choose(170,8)) + 8*log(1-psi) - 170*log(2-psi))
}

ml.psi <- maxLik2(loglik = loglik.psi, start = 0.9)

plot(ml.psi) +
  labs(title = "Second order approximation",
       subtitle = "n = 170, t = 8",
       x = expression(psi))
```

The plot shows that the MLE is extremely well approximated by the second order Taylor series, so the normal distribution is appropriate for the large sample distribution of our MLE.

Now, we test
$$
H_0 : \psi = 0.30\qquad \text{vs} \qquad H_1 : \psi > 0.30$$

The likelihood ratio test statistic is based on
$$
W = 2\ln \Lambda  \quad \text{where} \quad \Lambda = \frac{L(\hat \psi_0^\text{mle})}{L(\psi_0^\text{null})}
$$


$$
W = 2\left[\ell(\hat\psi_0^\text{mle}) - \ell(\psi_0^\text{null})\right].
$$
By substituting $\hat\psi_0^\text{mle} = \frac{154}{162}$ and $\psi_0^\text{null} = 0.3$ into our log likelihood function from, we get:

\begin{align*}
W &= 2 \Bigg[ \left( \ln\binom{170}{8} + 8\ln\left(1-\frac{154}{162}\right) -170\ln\left(2-\frac{154}{162}\right)\right) -\left(\ln\binom{170}{8} + 8\ln(1-0.3) -170\ln(2-0.3)\right)\Bigg] \\
   &= 121.6012
\end{align*}
Under $H_0$, the test statistic satisfies $W \sim \chi^2_1$ as we are only testing one parameter. Thus the P-value can be calculated as

$$
\text{P-value} = P(\chi^2_1 \ge W) = 1 - P(\chi_1^2<121.602)=2\times10^{-28}\approx0
$$
```{r label = "pval", eval = TRUE, echo = FALSE}
W <- 2*((8*log(8 / (162 * 0.7)) - 170*log(170/ (162* 1.7)))) 
p_val <- pchisq(W, df = 1, lower.tail = FALSE)
```

An alternative to calculate the P value is to calculate an empirical value by generating a large number of samples ($B=1000$) under the null hypothesis and calculating the proportion $\geq 121.602$.

```{r label = "empirical pval", eval = TRUE, echo = FALSE}
psi0 = 0.3
pi0 = (1 - psi0) / (2 - psi0)

# observed LRT statistic
W = 2*(loglik.psi(psi_mle) - loglik.psi(psi0))

w = numeric(B)

# Bootstrap
for(b in 1:B){

  Tb = rbinom(1,n,pi0)

  psi_mle = (n - 2*Tb)/(n - Tb)

  log_hat = loglik.psi(psi_mle)
  log_0   = loglik.psi(psi0)

  w[b] = 2*(log_hat - log_0)
}

# empirical p-value
p_val_b = mean(w >= W)
```

```{r label = "simulated Ws under null histogram", eval = TRUE, echo = FALSE, fig.height=4, fig.align='center'}
hist(w, breaks = 20, 
     xlab = "W", 
     ylab = "Frequency", 
     main = "Histogram of simulated W* values under null", 
     col = "lightblue",
     xlim = c(-60, 125))
```

The bootstrap method also gives us a p-value of 0.

## Bayesian Inference

In a Bayesian framework, we specify a prior distribution for $\pi$. A common choice for the binomial model is the Beta prior: $$g(\pi) = Beta(a, b)$$
We chose a skeptical prior that reflects uncertainty about whether the vaccine is effective. This allows the data to play a larger role in determining the posterior distribution and helps prevent incorrectly concluding that the vaccine is effective when it may not be.

Coming up with two prior beliefs about $\psi$:

\begin{enumerate}
  \item For $\psi \ge 0.30$, we have
  $$
  \frac{1 - 2\pi}{1 - \pi} \ge 0.30.
  $$
  Multiply both sides by $1-\pi$:
  $$
  1 - 2\pi \ge 0.30(1-\pi).
  $$
  Expanding gives
  $$
  1 - 2\pi \ge 0.30 - 0.30\pi.
  $$
  Rearranging,
  $$
  0.70 \ge 1.70\pi,
  $$
  so
  $$
  \pi \le \frac{0.70}{1.70} = \frac{7}{17}.
  $$
  Therefore,
  $$
  P\left(\pi \le \frac{7}{17}\right) = 0.05.
  $$

  \item For $\psi \ge 0$, we have
  $$
  \frac{1 - 2\pi}{1 - \pi} \ge 0.
  $$
  Since $1-\pi > 0$ for $0 < \pi < 1$, this is equivalent to
  $$
  1 - 2\pi \ge 0.
  $$
  Hence,
  $$
  \pi \le \frac{1}{2}.
  $$
  Therefore,
  $$
  P\left(\pi \le \frac{1}{2}\right) = 0.50.
  $$
\end{enumerate}

So we want a $\text{Beta}(a, b)$ prior for $\pi$ where the median is $\frac{1}{2}$ and the 5th percentile is $\frac{7}{17}$, or $P(\pi \le \frac{7}{17}) = 0.05$. Using the `beta.select` function from the `LearnBayes` library to find the beta prior, we obtain $a = 43.03$ and $b= 43.03$.

```{r label = "beta coeffs", eval=TRUE, echo=FALSE}
b_coeff = beta.select(quantile1 = list(p = 0.05, x = 7/17),
                      quantile2 = list(p=0.5, x = 0.5))
```

With the skeptical $Beta(43.03, 43.03)$, we have the beta prior for $\pi$:

```{r label = "beta prior plot", eval = TRUE, echo=FALSE, fig.height=4, fig.align='center'}
x_values = seq(0, 1, length.out = 1000)
alpha = b_coeff[1]
beta = b_coeff[2]
density_values = dbeta(x_values, shape1 = alpha, shape2 = beta)

plot(x_values, density_values, type = 'l',
     xlab = expression(pi), ylab = "PDF",
     main = paste("Beta(", alpha, ",", beta, ") prior for vaccine data", sep = ""))
```

The Beta distribution is conjugate to the binomial and so the posterior must also be a Beta distribution.

\begin{align*}
f(T\mid\pi) &\sim Binom(170, \pi) \text{ as the data model,} \\
g(\pi) &= Beta(43.04, 43.04) \text{ as the prior} \\
h(\pi\mid t) &= Beta(43.04 + t, 170 + 43.04 - t) \text{ as the posterior} \\
\text{where t = 8, } h(\pi\mid 8) &= Beta(43.04 + 8, 170 + 43.04 - 8) \\
\text{thus } h(\pi \mid 8) &= Beta(51.04, 205.04)
\end{align*}

```{r label = "prior vs post plot", eval = TRUE, echo = FALSE, fig.height=4, fig.align='center'}
x_values <- seq(0, 1, length.out = 1000)
n = 170
t = 8

# posterior parameters
alpha_post <- alpha + t
beta_post  <- beta + (n - t)

# densities
prior_density <- dbeta(x_values, shape1 = alpha, shape2 = beta)
post_density  <- dbeta(x_values, shape1 = alpha_post, shape2 = beta_post)

plot_data <- data.frame(
  pi = rep(x_values, 2),
  density = c(prior_density, post_density),
  distribution = rep(c("prior", "posterior"), each = length(x_values))
)

# plot
ggplot(plot_data, aes(x = pi, y = density, color = distribution)) +
  geom_line(linewidth = 0.5) +
  labs(
    title = "Beta prior and posterior for vaccine data",
    x = expression(pi),
    y = "PDF",
    color = "distribution"
  ) +
  theme_minimal()
```

```{r label = "posterior median", eval = TRUE, echo = FALSE}
posterior_median_pi = qbeta(0.5, shape1 = 51.04, shape2 = 205.04)
cat("Posterior Median: ", round(posterior_median_pi, 4))
cat("95% Confidence interval for pi (equal tails): (15.28, 25.03)")
```

```{r label = "posterior distribution", eval = TRUE, echo = FALSE, results='hide'}
qbeta(p = c(0.025, 0.975), shape1 = 51.04, shape2 = 205.04)
```

There is a 95% probability that the true probability of infection for a vaccinated individual lies in the range 0.1528 to 0.2503.

We can also find the highest posterior density interval (HPDI) for $\pi$ using the `HDInterval` library.

```{r label = "hdi interval", eval = TRUE, echo = FALSE}
posterior_interval = hdi(qbeta, credMass=0.95, shape1 = 51.04,
shape2 = 205.04)

cat("95% HPDI: ", round(posterior_interval, 4))
```

\newpage

Now that we have the posterior median and intervals for $\pi$, we want to transform it back in terms of $\psi$ in order to keep going with the analysis. The posterior median of $\pi$ is 0.1985, so $P(\pi < 0.1985 \mid t) = 0.5$.

Recall that $\psi = \frac{1-2\pi}{1-\pi}$.
\begin{align*}
& P(\frac{1-\psi}{2-\psi} < 0.1985\mid t) = 0.5 \\
& P(\psi > 0.7523\mid t) = 0.5
\end{align*}

This means that the posterior median of $\psi$ is 0.7523.

Getting the 95% equal tail interval for $\psi$:
\begin{align*}
&0.1528 \le \pi \le 0.2503 \\
&0.1528 \le \frac{1-\psi}{2-\psi} \le 0.2503
\end{align*}

\textbf{Lower bound}:
\begin{align*}
0.1528 &\le \frac{1-\psi}{2-\psi} \\
\psi &\le 0.8196
\end{align*}
\textbf{Upper bound} (right inequality):
\begin{align*}
\frac{1-\psi}{2-\psi} &\le 0.2503 \\
\psi &\ge 0.6661
\end{align*}
Thus, the 95\% equal tail CI for $\psi$: $0.6661 \le \psi \le 0.8196$.


Following similar steps for the HPDI: $0.6690\le \psi \le 0.8218$.

### Calculating the Bayesian P-value

The Bayesian analog of the P-value is to calculate the posterior probability of the null hypothesis.

To calculate the Bayesian P-value, generally,
\begin{align*}
\text{Bayes P-value} &: P(H_0 \text{ is TRUE }\mid X) \\
&= P(\psi \le \psi_0^{(null)}\mid X=x)
\end{align*}

In this scenario, $P(\psi \le 0.30 \mid 8)$, Therefore $P(\frac{1-2\pi}{1-\pi} \le 0.3 \mid 8)$ = $P(\pi \ge \frac{0.7}{1.7}\mid 8)=1.939\times10^{-13}$

```{r label = "bayesian pval", eval = TRUE, echo = FALSE, results='hide'}
pbeta(0.7/1.7, shape1 = 51.04, shape2 = 205.04, lower.tail = FALSE)
```

# Results

\begin{table}[htbp]
\centering
\caption{Bayesian Posterior Summaries}
\begin{tabular}{lcc}
\toprule
Quantity & $\pi$ & Vaccine Efficacy ($\psi$) \\
\midrule
Posterior Median & 0.1985 & 75.2\% \\
95\% Equal-Tail CI & [0.1528, 0.2503] & [66.6\%, 82\%] \\
95\% HPDI & [0.1513, 0.2486] & [66.9\%, 82.2\%] \\
\bottomrule
\end{tabular}
\end{table}

The posterior probability that the vaccine does not meet the FDA's 30% efficacy requirement is:

$$P(\psi \le 0.30\mid t=8) = P\biggr(\pi \ge \frac{0.7}{1.7} \mid t=8\biggr) = 1.939\times10^{-13}$$
Thus, there is significant evidence that the vaccine exceeds the 30% efficacy requirement set by the FDA.

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

\newpage

# Appendix
## Code

Initializing variables
```{r ref.label = "vars", eval=FALSE}

```

Contingency table
```{r ref.label = "table1", eval=FALSE}

```

Proportions histogram
```{r ref.label = "histogram", eval=FALSE}

```

Large sample confidence interval    
```{r ref.label = "confidence interval", eval=FALSE}

```

Second order taylor approximation plot
```{r ref.label = "Second order approximation plot", eval=FALSE}

```

Bootsrap percentile interval
```{r ref.label = "bootstrap", eval=FALSE}

```

P value
```{r ref.label = "pval", eval=FALSE}

```

Empirical P value
```{r ref.label = "empirical pval", eval=FALSE}

```

Simulated Ws under null histogram
```{r ref.label = "simulated Ws under null histogram", eval=FALSE}

```

Producing alpha and beta for beta distribution.
```{r ref.label = "beta coeffs", eval = FALSE}

```

Plotting $beta(\alpha, \beta)$ prior distribution
```{r ref.label = "beta prior plot", eval = FALSE}

```

Plotting beta prior and posterior distributions
```{r ref.label="prior vs post plot", eval=FALSE}

```

Calculating posterior median
```{r ref.label="posterior median", eval = FALSE}

```

Getting posterior distribution
```{r ref.label = "posterior distribution", eval = FALSE}

```

HDI interval
```{r ref.label = "hdi interval", eval = FALSE}

```

Bayesian p-value
```{r ref.label="bayesian pval", eval = FALSE}

```

## Proofs
If applicable, include detailed mathematical derivations or additional theoretical explanations.

### TODO: PROVE THAT THE DISTRIBUTION OF T CAN BE APPROXIMATED BY A BINOMIAL DISTRIBUTION
